Sustainable Pathways for Irish homes in changing TIMES

### 2021-07-23 manuscript.tex
Fix up the reference from mendely

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

\documentclass[authoryear,preprint,12pt]{elsarticle}
%\documentclass[10pt,final,journal,compsoc,a4paper]{IEEEtran}
%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{pgf-pie}
\usepackage{pgfplots}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{anyfontsize}
\usepackage{pgfplots}
\usepackage{comment}
\usepackage{booktabs}
\usepackage{rotating}
\usepackage{siunitx}
\usepackage{eurosym}
\usepackage{multirow}
\usepackage{xcolor} 
\usepackage{colortbl}
\usepackage{subcaption}

\DeclareSIUnit{\EUR}{\text{\euro}}
 \pgfplotsset{compat=1.17}
%\usepackage[table]{xcolor}
%\usepackage{cite}
%\usepackage{harvard}
%\usepackage{biblatex}
\addbibresource{references.bib} %Imports bibliography file

%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.

@@

\author[inst1,inst2]{Jason Mc Guire\footnote{Contact: j.mcguire@ucc.ie}}

\affiliation[inst1]{organization={Energy Policy and Modelling Group, MaREI Centre},%Department and Organization
            addressline={Environmental Research Institute}, 
            city={Cork},
            country={Ireland}}

@@

\begin{abstract}
%% Text of abstract
%Ireland's residential sector has particularly underachieved in previous climate targets. Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector. The disadvantageous starting position has impeded residential decarbonisation progress thus far. \par
%Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy. \par
%The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
\par Through historical settlement patterns, thermally inefficient building fabrics, high fuel poverty, and the social aspects of solid fuel heating, the Irish residential sector has proved hard to decarbonise. Despite this, Ireland has set one of the most ambitious climate targets into law, a 51 \% greenhouse gas reduction in 2030, compared to 2018. There is a need for a new state of the art model to provide insights on the optimal pathway for the energy system. This paper reviews residential and energy system models and describes the methodology applied in the residential sector in the newly developed TIMES Ireland Model (TIM). The paper includes some results to provide a discussion on some of the insights which need to be further investigated in the residential sector.

@@

%%Graphical abstract
%%\begin{graphicalabstract}
%%\includegraphics{grabs}
%%\end{graphicalabstract}

%%Research highlights
%%\begin{highlights}
%%\item Research highlight 1
%%\item Research highlight 2
%%\end{highlights}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Energy Systems Optimisation Model \sep TIMES \sep Residential decarbonisation \sep Model description 
%% PACS codes here, in the form: \PACS code \sep code
%%\PACS 0000 \sep 1111
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
%% \MSC 0000 \sep 1111
\end{keyword}

@@

Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways and interpreting the results. Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy, and distinctive residential sector. While also examining how researchers have previously used bottom-up residential models and Energy System Optimization Models (ESOMs) to analyse the residential sector.
Section \ref{LitR} describes the methodology used to develop the residential sector. The treatment of input data is described in the model,  while the calculations used to develop the base year residential dwelling stock, and cost, saving and demand estimates are outlined, as is the process of applying constraints in the model.  
Section \ref{Results} outlines the results obtained from the six selected scenarios. Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models. Lastly, section \ref{Conclusion} draws some conclusions on the model, results and discussion on future work.

@@

Energy System Optimisation Models (ESOMs) can be used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOMs cover all energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.

\par For example, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK using UK TIMES Model (UKTM). As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. Li set out to incorporate a Discrete Choice Model (DCM) for TIMES to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices. While Li's research has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights through more surveys and analysis.

Leibowicz et al. developed an expanded version of the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.} that integrates electricity generation and storage, residential building energy service demands, and end-use technologies \cite{Leibowicz2018OptimalServices}.Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas. Austin was chosen because of the high granularity of available data \cite{EnergyInc.}. Leibowicz et al. results find net present costs in the \textit{``Policy scenarios"} ( emissions declining linearly toward 10\% of their current level by 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ). However, the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock. Leibowicz et al. also found electrification to be the dominant strategy for cost-effectively decarbonizing urban residential dwellings. These insights show the strength of an ESOM, but not all costs are taken into account in Leibowicz study, the capital costs associated with more thermally efficient buildings are not included.

Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark. The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland, and Germany to cover import/export. Exogenous user constraints applied to the model, based on the current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030, and individual oil boilers are phased out by 2030. The results find that when investments in individual heat pumps are allowed, air-to-water heat pumps are installed in non-district heating areas. This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and orangeucing system costs, fuel consumption, and $CO_2$ emissions.
Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is an reduced need for peak/reserve capacity investments. The Balmorel model has provided insights, but it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies. Nevertheless, the insights are valuable and the results can be soft-linked to other models for deeper analysis.

Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at the household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island. 
Two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach. Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be storage in communal batteries. When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery. While the battery location was exploorange, the PV system location was not, further investigation into the optimal PV system location could provide further insights. The household occupancy behaviour with and without household batteries on the PV system could alter the optimal pathway.

@@

% NOT INCLUDE HAVE ENOUGH Petrovic and Karlsson \cite{Petrovic2016ResidentialSystem}  examined a new approached to represent residential heat pumps in Denmark

@@

ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2021AalborgEnergy-System}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.

@@

Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector. This was the first comprehensive analysis of the residential sector, and their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}. In 2000, Clinch and Healy developed and published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating. A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts. Under the predicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
The same model was used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component
into the computer model. The modelling process involved developing a computer model of the energy performance of the existing dwelling stock. It was noted that the energy consumption predicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions. 
The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.

@@

The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}. ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates the annual energy consumption to simulate the potential for improved energy efficiency gains within the existing retrofit programs. Mac Uidhir et al. specifically examined the Better Energy Homes (BEH) scheme. The five most common retrofit combinations are simulated for each of the 9 archetypes ( 3 energy rating types and 3 building types ). The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when comparing the baseline scenario. This shows the the type of retrofit and type of house can have a large affect on the energy saved. 
Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is consideorange negligible.

Finally, it is worth mentioning SEAI’s National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system. Some of the main models used in the NEMF include:
\begin{itemize}
\item {Core Structural Model of the Irish Economy (COSMO)  is a macro-economic model used by the Economic and Social Research Institute (ESRI) to create a baseline forecast of energy consumption by sector} \cite{Bergin2017COSMO:Ireland}
\item{PLEXOS is a power systems modeling tool used for electricity modeling and planning}\cite{PLEXOSExemplar}
\item{Energy-efficiency policy analysis tool informs the energy efficiency estimates} \cite{Scheer2015UnlockingTob}
\begin{itemize}
\item{Consumer Choice Model determines the attractiveness of energy efficiency finance options}\cite{SustainableEnergyAuthorityofIreland2017BehaviouralSector}
\end{itemize}
\item{BioHEAT is a techno-economic simulation model of bioenergy and heat, used by SEAI to support policy decision in Ireland.}\cite{Durusut2018BioHEAT:Sectors}
\end{itemize}

@@

Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings, and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx. 85\% of residential buildings are 2005 or older. Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil, accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.

In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential Greenhouse Gas (GHG) emissions. GHG emissions from coal in the residential sector also declined to 6.9\% in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. However, the decline in peat and coal has been offset by a more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. While, gas availability is limited to some urban areas and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. 
The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}. Preliminary results from the Environmental Protection Agency (EPA) have indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.

@@

\begin{figure}[!htbp]
    \begin{tikzpicture}[every node/.style={align=center}, pin distance=17mm]
    \pie[cloud,text=inside,radius = 2.4]{34/Agriculture,7/Other,59/Energy-related}
    \draw [->, line width=4pt] (5.4,-0.3) -- (6.5,-.3);
    \hspace{8.5cm}
    \pie[rotate =30,text=inside,radius=1.95,color={black!30,orange!70,yellow!40}]{40/Transport,33/Heat,27/Electricity}
    \end{tikzpicture}
    \caption{Ireland 2018 GHG Share \cite{Howley2005ENERGY-RELATEDIreland}}
    \label{fig:Ireland2018GHG}
\end{figure}

@@

%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
\newpage

@@

>>>>

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references
%\usepackage{cite}
%\usepackage{harvard}
%\usepackage{biblatex}
%\usepackage[backend=biber]{biblatex}
% \addbibresource{references.bib} %Imports bibliography file

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
%\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[preprint,12pt]{elsarticle}
%\documentclass[10pt,final,journal,compsoc,a4paper]{IEEEtran}
%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%%\documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%%\documentclass[final,3p,times,authoryear]{elsarticle}
%%\documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{pgf-pie}
\usepackage{pgfplots}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{anyfontsize}
\usepackage{pgfplots}
\usepackage{comment}
\usepackage{booktabs}
\usepackage{rotating}
\usepackage{siunitx}
\usepackage{eurosym}
\usepackage{multirow}
\usepackage{xcolor} 
\usepackage{colortbl}
\usepackage{subcaption}
\usepackage{blindtext}
\usepackage{lineno,hyperref}
\modulolinenumbers[5]
\DeclareSIUnit{\EUR}{\text{\euro}}
 \pgfplotsset{compat=1.17}
%\usepackage[table]{xcolor}
\usepackage{biblatex}

%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}
\addbibresource{references.bib}
%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.

@@

\author[inst1,inst2]{Jason Mc Guire \corref{j.mcguire@ucc.ie}}
\cortext[j.mcguire@ucc.ie]{j.mcguire@ucc.ie}
%footnote{E-mail: \url{j.mcguire@ucc.ie}}}

%\footnote{Contact: j.mcguire@ucc.ie}}
%\thanks{Email: \url{j.mcguire@ucc.ie}}}
\affiliation[inst1]{organization={Energy Policy and Modelling Group, MaREI Centre},%Department and Organization
            addressline={Environmental Research Institute}, 
            city={Cork},
            country={Ireland}}

@@

\begin{abstract}
%% Text of abstract
%Ireland's residential sector has particularly underachieved in previous climate targets. Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector. The disadvantageous starting position has impeded residential decarbonisation progress thus far. \par
%Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy. \par
%The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
\par The Irish residential sector has proved hard to decarbonise, yet recently Ireland has set one of the most ambitious climate targets into law, a 51 \% greenhouse gas reduction in 2030, compared to 2018. There is a need for a new state of the art model to provide insights on the optimal decarbonisation pathway for the Irish energy system. This paper reviews residential and energy system models and describes the methodology applied to the development of the residential sector in the new TIMES Ireland Model (TIM). The paper includes some results to provide a discussion on some of the insights and highlights the insights which may need to be further investigated in the residential sector.

@@

\begin{keyword}

Energy Systems Optimisation Model \sep TIMES \sep Residential decarbonisation \sep Model description

\end{keyword}

@@

Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways and interpreting the results. Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy, and distinctive residential sector. While also examining how researchers have previously used bottom-up residential models and Energy System Optimization Models (ESOMs) to analyse the residential sector.
Section \ref{LitR} describes the methodology used to develop the residential sector. The treatment of input data in the model is described, the calculations are explained, and assumptions and constraints are highlighted. 
Section \ref{Results} outlines the six sceanrios used and displays the energy system and residential results. Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models. Lastly, section \ref{Conclusion} draws some conclusions on the model, results and discussion on future work.

@@

ESOMs have been used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOMs currently used include MESSAGE \cite{Messner1995Model-BasedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOMs cover all energy-related sectors, they can also be used to specifically explore certain economic sector in higher granularity.

TIMES has been extensively used in both the UK and Ireland, and there are several examples using TIMES to focus on the residential sector. TIMES assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. With this in mind, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices, this identifies perfect foresight as a weakness in TIMES. While Li's research has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights through more surveys and analysis.
 \par Many sectors are increasingly reliant on electricity generation as lower carbon intensity electricity is considered a cornerstone in optimal decarbonisation. Using the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.}
Leibowicz et al.\cite{Leibowicz2018OptimalServices} explores the optimal decarbonisation of the residential sector by examining electricity generation and storage strategies. The results find net present costs in the \textit{``Policy scenarios"} ( $CO_2$ declines linearly by 90\% in 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ). However, the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock, highlighting the importance of energy building regulation as thermally efficient building stock can reduce the decarbonisation cost. 
District Heating is a proven low carbon heating technology, however it is almost non-existant and has not been included in previous ESOM in Ireland. Denmark are leading experts in this area and Denmark's district heating system has been comprehensively examined. One study by 
Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} used the Balmorel model to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark. Exogenous user constraints applied to the model, based on the current Danish policy. The results find that air-to-water heat pumps are installed in non-district heating areas, while the system benefits of adding individual heat storage to the heat pumps are only moderate, the main benefit is a reduced need for peak/reserve capacity investments. This identifies individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments in areas where district heating is not available.
Another technology which is proven but under utilised in Ireland, is photovoltaic (PV) solar panels. While Ireland's policy and building regulations have encouraged the uptake of PVs, however the market share of PV systems remains low, and as the cost continues to fall they can alter optimal decarbonisation pathways.   
Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model to explore the feasibility of PV systems installed at the household level in combination with batteries in Samsø.
Two approaches are explored for the location of the batteries in the PV system; communal battery system or an individual residential battery approach. Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be storage in communal batteries. When identical communal and individual battery capacities where compared, it was found  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery. While the battery location was explored, the PV system location was not. This is a great example of the strength of energy system models, whereby the residential PV systems affect wind energy investment. However, a weakness in the model is the ability to comprehend household occupancy behaviour with and without household batteries on the PV system.

@@

@@

ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2009DevelopingUniversity.}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.

@@

Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector. This was the first comprehensive analysis of the Irish residential sector, and their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}. In 2000, Clinch and Healy published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating. A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts. Under the predicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
The same model was used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component into the computer model. The modelling process involved developing a computer model of the energy performance of the existing dwelling stock. It was noted that the energy consumption predicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions. 
The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.

@@

The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}. ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates the annual energy consumption to simulate the potential for improved energy efficiency gains within the existing retrofit programs. Mac Uidhir et al. specifically examined the Better Energy Homes (BEH) scheme. The five most common retrofit combinations are simulated for each of the 9 archetypes ( 3 energy rating types and 3 building types ). The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when comparing the baseline scenario. This shows the the type of retrofit and type of house can have a large affect on the energy saved. 
Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is negligible.

Finally, it is worth mentioning SEAI’s National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system. Some of the main models used in the NEMF include:
\begin{itemize}
\item {Core Structural Model of the Irish Economy (COSMO)  is a macro-economic model used by the Economic and Social Research Institute (ESRI) to create a baseline forecast of energy consumption by sector} \cite{Bergin2017COSMO:Ireland}
\item{PLEXOS is a power systems modeling tool used for electricity modeling and planning}\cite{2021PLEXOSExemplar}
\item{Energy-efficiency policy analysis tool informs the energy efficiency estimates} \cite{Scheer2015UnlockingOpportunity}
\begin{itemize}
\item{Consumer Choice Model determines the attractiveness of energy efficiency finance options}\cite{SustainableEnergyAuthorityofIreland2017BehaviouralSector}
\end{itemize}
\item{BioHEAT is a techno-economic simulation model of bioenergy and heat, used by SEAI to support policy decision in Ireland.}\cite{Durusut2018BioHEAT:Sectors}
\end{itemize}

@@

Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings, and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{Eurostat2021DistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{Eurostat2021ShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousing2021Gov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx. 85\% of residential buildings are 2005 or older. Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil, accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020EnergySEAI}.

In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential Greenhouse Gas (GHG) emissions. GHG emissions from coal in the residential sector also declined to 6.9\% in 2018 \cite{Howley2005Energy-RelatedIreland.}. However, the decline in peat and coal has been offset by a more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005Energy-RelatedIreland.}. While, gas availability is limited to some urban areas and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005Energy-RelatedIreland.}. 
The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}. Preliminary results from the Environmental Protection Agency (EPA) have indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.

Space heating is the largest residential energy service, which accounted for 61.8\% of final energy usage in 2018. Other residential energy services and their share of final energy are shown in Fig.\ref{fig:Ireland2018ServiceDemands}. As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline.

@@

\begin{figure}[!htbp]
    \begin{tikzpicture}[every node/.style={align=center}, pin distance=17mm]
    \pie[cloud,text=inside,radius = 2.4]{34/Agriculture,7/Other,59/Energy-related}
    \draw [->, line width=4pt] (5.4,-0.3) -- (6.5,-.3);
    \hspace{8.5cm}
    \pie[rotate =30,text=inside,radius=1.95,color={black!30,orange!70,yellow!40}]{40/Transport,33/Heat,27/Electricity}
    \end{tikzpicture}
    \caption{Ireland 2018 GHG Share \cite{Howley2005Energy-RelatedIreland.}}
    \label{fig:Ireland2018GHG}
\end{figure}

@@

@@

%\bibliographystyle{elsarticle-harv}

%\bibliography{references.bib}
